---
title: "Insurance Coverage and Psychiatric Drug Exposure"
author:
  - name: Sunny Lee
    corresponding: true
    email: snylee@uchicago.edu
    affiliations:
      - name: University of Chicago
        department: Department of Comparative Human Development
        address: 1101 E 58th St
        city: Chicago
        region: IL
        postal-code: 60637
abstract: "This study examines the relationship between insurance coverage and psychiatric drug exposure using data from the 2022 Medical Expenditure Panel Survey (MEPS). Given the increasing reliance on psychotropic medication for mental health treatment, understanding how insurance coverage influences access to these medications is crucial. We hypothesized that individuals with better insurance coverage would have higher psychiatric drug exposure. The primary findings from chi-square tests show that individuals with private or public insurance (Medicare or other) are almost twice as likely to be exposed to psychiatric drug prescriptions compared to their uninsured counterparts. This relationship remains significant across all income levels, except for the 'Near Poor' category, likely due to limitations in sample size rather than an actual lack of statistical significance. Logistic regression results reinforce these findings, showing that regardless of age or Medicare eligibility, insured individuals have approximately twice the odds of being prescribed psychiatric medication compared to those without insurance. These findings emphasize the critical role of insurance coverage in mitigating disparities in mental health treatment and highlight the need for policy interventions to improve psychiatric medication access for economically vulnerable populations."
keywords: [MEPS, Medical Expenditure Panel Survey, Mental Health, Poverty, Prescribed Medicine, Psychiatric Drug, Drug Expenditure]
author-note:
disclosures:
  conflict of interest: The author has no conflict of interest to declare.
format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
bibliography: bibliography.bib
execute:
  warning: false  # Suppresses warnings
---

# Introduction
Mental health disorders pose a significant global challenge, with more than half of individuals suffering from mental illnesses lacking access to proper care [@pateletal_2010_reducing]. This treatment gap is widespread across nations but disproportionately affects individuals in lower-income brackets. The relationship between poverty, mental health, and social risks including stress, stigma, and unemployment has been extensively studied, revealing a cyclical interaction in which mental illness can lead to financial hardship while economic instability exacerbates psychiatric conditions [@kniftoninglis__poverty]. @marbinetal_2022_perspectives has shown that income inequality also has been linked to heightened mental disorder problems in developed countries. Despite the heightened risk of psychiatric disorders among socioeconomically disadvantaged populations, significant barriers including financial constraints, lack of insurance, and limited healthcare access impede treatment utilization [@dongetal_2021_comparison].

The treatment of mental disorders have increasingly geared toward reliance on psychotropic drugs, given its easier accessibility compared to in-patient care or regular office visits[@donohuefrank_2007_estimating]. One of the primary obstacles to mental healthcare access is the cost of psychiatric medications, which constitute a substantial portion of healthcare expenditures [@dongetal_2021_comparison]. The rising cost of psychotropic drugs has placed an increasing financial burden on low-income families, leading to a higher prevalence of unmet healthcare needs [@sarmaetal_2007_influence]. Insurance coverage plays a crucial role in mitigating these disparities, as research has shown that expanding coverage improves access to prescription medications. For example, a 2018 study in Oregon found that Medicaid coverage reduced the prevalence of untreated depression by over 60%, while increasing the use of antidepressants and other psychiatric medications [@baickeretal_2018_effect]. Similarly, studies have found that cost-sharing declines psychotropic drug usage, even in cases where usage is necessary for patients[@fungetal_2013_adverse]. These findings underscore the importance of financial barriers in psychiatric medication adherence, highlighting the role of insurance policies in reducing disparities in mental healthcare access.

Given the critical role of insurance coverage in determining psychiatric drug access, this study examines the following research question: **What is the relationship between insurance coverage and exposure to psychiatric drugs?** Based on prior research, the study hypothesizes that individuals with comprehensive insurance coverage will have greater exposure to psychiatric medications compared to those with limited or no insurance.

This study utilizes data from the Medical Expenditure Panel Survey (MEPS) to investigate the association between insurance coverage and psychiatric drug exposure, obtained from the Prescribed Medicine File. The primary outcome variable, psychiatric drug exposure, is defined as a binary indicator of whether an individual was prescribed psychiatric medication, as reported by the household and confirmed by a pharmacy with individual consent. The key predictor variable, insurance coverage, is categorized into private insurance, public insurance (Medicaid, Medicare), and uninsured status. Several covariates that may influence psychiatric drug exposure are also considered, including:

Number of psychiatric diagnoses
**Demographic characteristics**: Age, race/ethnicity, and sex
**Socioeconomic status**: Poverty/income level. 

# Methods

## Data Source
This study utilizes data from the **Medical Expenditure Panel Survey (MEPS)**, a nationally representative survey that collects information on healthcare utilization, expenditures, and insurance coverage in the United States. Specifically, the **2022 MEPS Full-Year Consolidated File (HC 243), Medical Conditions File (HC 241), and Prescribed Medicine File (HC 239a)** were used. The **household component** of MEPS provides person-level data, while the **medical conditions and prescribed medicine files** include condition- and prescription-level data, respectively. The **2022 dataset** included **21,747 individuals** [^1]. Participants included in this study were individuals aged **18 years and older**.

[^1]: [@__medical]

## Data Preparation
### Variable Construction
- **Psychiatric Drug Exposure:** A binary outcome variable indicating whether an individual was prescribed and purchased a psychiatric drug. This was derived from the **HC 239a prescribed medicine file**, where only medications classified under psychiatric therapeutic classes were retained.
- **Psychiatric Diagnoses:** Conditions were filtered using **ICD codes** from the **HC 241 medical conditions file**, and only psychiatric conditions were retained.
- **Insurance Coverage:** The original MEPS insurance categories were **recoded** as follows:
  - **<65 Any Private and <65 Public Only → <65 Any Private/Public**
    - **65+ Medicare and Private, 65+ Medicare and Other Public Only → 65+ Medicare and Any Private/Public**
      - **65+ No Medicare and Any Public/Private → 65+ No Medicare and Any Private/Public**
Age was both left as a continuous variable and categorized as **<65 years** and **65+ years** to align with Medicare eligibility thresholds.

## Key Variables
- **Outcome Variable:** Psychiatric drug exposure (Yes/No)
- **Independent Variable:** Insurance coverage status, categorized as:
  - <65 Uninsured
  - <65 Any Private/Public
  - 65+ without Medicare
  - 65+ with Medicare
- **Covariates:**
  - Age
  - Sex
  - Race/Ethnicity
  - Income Level

## Statistical Analysis

### Descriptive Analysis
Descriptive statistics were computed for demographic and socioeconomic characteristics, stratified by psychiatric drug exposure status. **Chi-square tests** examined bivariate relationships between psychiatric drug exposure and key variables.

### Logistic Regression Analysis
Separate **logistic regression models** were performed for each age group to assess the association between **insurance coverage and psychiatric drug exposure**. Odds ratios (OR) with **95% confidence intervals (CIs)** were estimated. The reference groups were:
- **<65 Uninsured** for the <65 model
- **65+ without Medicare** for the 65+ model

All analyses were conducted using **R version 2024.12.0+467** with the `tidyverse`, `gtsummary`, `knitr`, `kable`, `papaja`, `broom`, and `flextable` packages.

# Results
```{r}
#| label: sourcing-data
#| include: false

library(readr) #Run for every new R session

#Source an R script that has imported ASCII file and data-cleaned, extracting only relevant variables
source ("data-wrangling/Loading and cleaning h243.R")

#Source an R script that merges full year consolidated and medical conditions file 
source ("data-wrangling/Merged h243 and h241.R")

#Source all 
source ("data-wrangling/Added h239a to data.R")
```

```{r}
#| label: load-packages
#| include: false 

#install.packages("gtsummary") #commented out the install function so it doesn't install everytime
#install.packages("flextable")
library(tidyverse) 
library(gtsummary) #Creates summary tables for descriptive statistics
library(knitr) #Report generation and formatting
library(kableExtra) #Better tables
library(scales) # Squishing scales (and other outlier handling)
library(psych) # Common analyses and visualizations in psychology
library(stats) # Basic statistics
library(lme4) # Mixed models
library(broom) # Tidy model output
library(papaja) # General APA friendly stuff, but old and not always quarto-friendly
# load flextable after papaja because they both have a theme_apa() function
# and papaja's one (for plots) isn't very reliable; we want flextable's (for tables) to override ("mask") it
library(flextable) # APA table 

# 3️⃣ Use `require()` to load packages (does not stop execution if missing)
if (!require(dplyr)) {
  install.packages(dplyr)  # Install if missing
  library(dplyr)  # Load the package #Data manipulation (included in tidyverse)
}

source("functions.R")
```

```{r}
#| label: bib-setup
#| include: false 

write_bib(c("ggplot2", "dplyr", "tidyverse", "gtsummary", "kableExtra", "papaja", "flextable"), "references.bib")

```

```{r}
#| label: analysis-preferences
#| include: false

# Seed for random number generation
set.seed(24)

# Set global chunk options
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed,
  fig.height = 6,
	fig.width = 7,
	message = FALSE,
	warning = FALSE,
	comment = ""
  )
```

## Descriptive Statistics

```{r}
#| label: descriptive-stat-data
#| include: false 

# Generate descriptive statistics by psychiatric drug exposure
table1 <- clean_data %>%
  group_by(exposure_psych_drugs) %>%
  summarise(
    `Age (Mean ± SD)` = paste0(round(mean(AGE31X, na.rm = TRUE), 2), 
                                " (±", round(sd(AGE31X, na.rm = TRUE), 2), ")"),
    `Male (%)` = paste0(round(mean(SEX == "Male", na.rm = TRUE) * 100, 2), "%"),
    `Female (%)` = paste0(round(mean(SEX == "Female", na.rm = TRUE) * 100, 2), "%"),
    `Hispanic (%)` = paste0(round(mean(RACETHX == "Hispanic", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic White (%)` = paste0(round(mean(RACETHX == "Non-Hispanic White", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic Black (%)` = paste0(round(mean(RACETHX == "Non-Hispanic Black", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic Asian (%)` = paste0(round(mean(RACETHX == "Non-Hispanic Asian", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic Other/Multiple (%)` = paste0(round(mean(RACETHX == "Non-Hispanic Other/Multiple", na.rm = TRUE) * 100, 2), "%"),
    `Poor/Negative (%)` = paste0(round(mean(POVCAT22 == "Poor/Negative", na.rm = TRUE) * 100, 2), "%"),
    `Near Poor (%)` = paste0(round(mean(POVCAT22 == "Near Poor", na.rm = TRUE) * 100, 2), "%"),
    `Low Income (%)` = paste0(round(mean(POVCAT22 == "Low Income", na.rm = TRUE) * 100, 2), "%"),
    `Middle Income (%)` = paste0(round(mean(POVCAT22 == "Middle Income", na.rm = TRUE) * 100, 2), "%"),
    `High Income (%)` = paste0(round(mean(POVCAT22 == "High Income", na.rm = TRUE) * 100, 2), "%")
  ) %>%
  pivot_longer(cols = -exposure_psych_drugs, names_to = "Variable", values_to = "Value") %>%
  pivot_wider(names_from = exposure_psych_drugs, values_from = Value)

# Rename columns for better readability
colnames(table1) <- c("Variable", "No Exposure (%)", "Exposed (%)")
```

```{r}
#| label: tbl-descriptive-stat
#| tbl-cap: "Descriptive statistics by exposure to psychiatric drugs"
#| apa-note: "Race and Income Level were all self-reported and categorized by MEPS standards"
#| echo: false

# Format the table with APA-style formatting
table1 %>%
  kable("html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  pack_rows("Age", 1, 1) %>%
  pack_rows("Sex", 2, 3) %>%
  pack_rows("Race/Ethnicity", 4, 8) %>%
  pack_rows("Income Level", 9, 13) %>%
  add_indent(c(2:3, 4:8, 9:13))  # Indent subcategories

```
@tbl-descriptive-stat presents descriptive statistics for the study population by psychiatric drug exposure status. Individuals exposed to psychiatric drugs were more likely to be 
**older**, **female**, and from **higher-income** groups.

```{r}
#| label: age-category
#| include: false

#Make Age a Categorical Variable 
clean_data <- clean_data %>%
  mutate(AGE_GROUP = case_when(
    (AGE31X <= (10 + 8)) ~ "0~18",
    (AGE31X >= (10 + 8)) & (AGE31X <= (20 + 5)) ~ "18-25",  # Arithmetic: 10+8, 20+5
    (AGE31X >= (20 + 6)) & (AGE31X <= (40 + 5)) ~ "26-45",
    (AGE31X >= (40 + 6)) & (AGE31X <= (60 + 4)) ~ "46-64",
    (AGE31X >= (60 + 5)) ~ "65+",
    TRUE ~ NA_character_  # Ensuring NA handling
  ) %>% as.factor())  # Assignment operator: = and %>%

```

## Chi-Square Test Results
```{r}
#| label: run-chi-square
#| include: false

# Run Chi-Square tests
sex_chi <- chisq.test(table(clean_data$SEX, clean_data$exposure_psych_drugs))
race_chi <- chisq.test(table(clean_data$RACETHX, clean_data$exposure_psych_drugs))
age_chi <- chisq.test(table(clean_data$AGE_GROUP, clean_data$exposure_psych_drugs))  # Use AGE_GROUP instead of Age
pov_chi <- chisq.test(table(clean_data$POVCAT22, clean_data$exposure_psych_drugs))
```

```{r}
#| label: chi-square-df
#| include: false
# Create a data frame with results
chi_square_bygroup <- data.frame(
  Variable = c("Sex", "Race/Ethnicity", "Age Group", "Poverty Level"),
  `X²` = round(c(sex_chi$statistic, race_chi$statistic, age_chi$statistic, pov_chi$statistic), 3),  # Round X² values to 3 decimal places
  df = c(sex_chi$parameter, race_chi$parameter, age_chi$parameter, pov_chi$parameter),
  p_value = c(sex_chi$p.value, race_chi$p.value, age_chi$p.value, pov_chi$p.value)
)

# Format p-values for APA style
chi_square_bygroup$p_value <- as.character(ifelse(chi_square_bygroup$p_value < .001, "< .001", round(chi_square_bygroup$p_value, 3)))
```

```{r}
#| label: tbl-chi-square
#| tbl-cap: "Relationship between Insurance Level and Exposure to Psychiatric Drugs based on Covariates"
#| echo: false

# Make APA-style Chi-Square Test Table
chi_square_table <- flextable(chi_square_bygroup) %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(align = "center", part = "body") %>%
  width(j = 2:4, width = 2) %>%  
  fontsize(size = 10, part = "all") %>%
  autofit() %>%
  italic(j = 4, part = "body")  # Italicize p-values (APA style)

# Print the table
chi_square_table
```
Chi-square tests indicated significant differences in **psychiatric drug exposure** across **`r chi_square_bygroup$Variable[1]`**, **`r chi_square_bygroup$Variable[2]`**, **`r chi_square_bygroup$Variable[3]`**, and **`r chi_square_bygroup$Variable[4]`** (@tbl-chi-square). 

The chi-square test for **`r chi_square_bygroup$Variable[1]`** was **X²**(`r sex_chi$parameter`) = `r round(sex_chi$statistic, 3)`, p = `r apa_p(sex_chi$p.value)`. This suggests a significant relationship between sex and psychiatric drug exposure. For **`r chi_square_bygroup$Variable[2]`**, was **X²**(`r race_chi$parameter`) = `r round(race_chi$statistic, 3)`, *p* = `r apa_p(race_chi$p.value)`, indicating a significant association between racial identity and exposure to psychiatric drugs.
For **`r chi_square_bygroup$Variable[3]`** was **X²**(`r age_chi$parameter`) = `r round(age_chi$statistic, 3)`, *p* = `r apa_p(age_chi$p.value)`. This suggests that psychiatric drug exposure significantly varies by age group.
For **`r chi_square_bygroup$Variable[4]`** was **X²**(`r pov_chi$parameter`) = `r round(pov_chi$statistic, 3)`, *p* = `r apa_p(pov_chi$p.value)`. This indicates a significant association between socioeconomic status and psychiatric drug exposure.

## Subgroup Chi-Square Analysis

```{r}
#| label: sub-group-chisq 
#| include: false 
# Make wide table
summary_table <- clean_data %>%
  group_by(POVCAT22, INSURC22, exposure_psych_drugs) %>%
  summarise(count = n(), .groups = "drop") %>% #Drop groups after summarizing 
  tidyr::pivot_wider(names_from = exposure_psych_drugs, values_from = count, values_fill = 0)

# Chi-square test within each income level and Fischer's exact for smaller values  
subgroup_chisq <- clean_data %>%
  group_by(POVCAT22) %>%
  summarise(
    p_value = tryCatch({
        if (any(table(INSURC22, exposure_psych_drugs) < 5)) {
          fisher.test(table(INSURC22, exposure_psych_drugs), simulate.p.value = TRUE)$p.value}          else {
          chisq.test(table(INSURC22, exposure_psych_drugs))$p.value
        }
      },
      error = function(e) NA ))  # If error occurs, return NA

# Calculate "n" values for individuals with "Yes" in exposure to psych drugs per income level
n_values <- clean_data %>%
  filter(exposure_psych_drugs == "Yes") %>%
  group_by(POVCAT22) %>%
  summarise(n = n(), .groups = "drop")
```

```{r}
#| label: sub-group-format
#| include: false 
# Merge the "n" values with chi-square results
subgroup_chisq <- subgroup_chisq %>%
  left_join(n_values, by = "POVCAT22")

# Format p-values for APA style
subgroup_chisq <- subgroup_chisq %>%
  mutate(p_value = ifelse(p_value < 0.001, "< .001", sprintf("%.3f", p_value)))

# Rename columns for APA format
subgroup_chisq <- subgroup_chisq %>%
  rename(
    "Income Level" = POVCAT22,
    "Chi-Square / Fisher's p-value" = p_value,
    "n (Exposed to Psych Drugs)" = n
  )
```

```{r}
#| label: tbl-subgroup-analysis
#| tbl-cap: "Relationship between Insurance Level and Exposure to Psychiatric Drugs based on Income Level"
#| apa-note: "Some subgroups didn't have enough representation due to the nature of MEPS data"
#| echo: false

# Create APA-style table with "n" values
table3 <- flextable(subgroup_chisq) %>%
  theme_vanilla() %>%  # APA-style theme
  bold(part = "header") %>%  # Make headers bold
  align(align = "center", part = "all") %>%  # Center align all content
  fontsize(size = 10, part = "all") %>%
  add_footer_lines("Note. Some subgroups didn't have enough representation due to the nature of MEPS data.") %>%
  width(j = 2, width = 2) %>%  # Adjust X² column width
  width(j = 3, width = 2) %>%  # Adjust df column width
  autofit()

#Print table
table3
```
Subgroup analyses stratified by income level showed that the association between insurance coverage and psychiatric drug exposure remained statistically significant in all income groups except for the Near Poor category (@tbl-subgroup-analysis).

For example, among individuals in the **`r subgroup_chisq[["Income Level"]][1]`** category, the chi-square test indicated a significant association `r get_chisq_p("Poor/Negative", subgroup_chisq)`, suggesting that psychiatric drug exposure differs by insurance coverage in this income level.
Similarly, in the **`r subgroup_chisq[["Income Level"]][2]`** category, the relationship between insurance level and psychiatric drug exposure remained significant `r get_chisq_p("Low Income", subgroup_chisq)`, reinforcing the role of insurance coverage in determining psychiatric medication access.
For the **`r subgroup_chisq[["Income Level"]][3]`** group, a statistically significant association was observed `r get_chisq_p("Middle Income", subgroup_chisq)`, indicating that psychiatric drug exposure varied across different insurance statuses.
Likewise, in the **`r subgroup_chisq[["Income Level"]][4]`** category, the chi-square test remained statistically significant `r get_chisq_p("High Income", subgroup_chisq)`, further supporting the notion that individuals with different insurance statuses experience varying levels of access to psychiatric medication.
In contrast, the **`r subgroup_chisq[["Income Level"]][5]`** category did not show a statistically significant association `r get_chisq_p("Near Poor", subgroup_chisq)`, implying that insurance coverage might not play a decisive role in psychiatric drug exposure within this income level.


```{r}
#| label: change-insurance-level
#| include: false 
clean_data <- clean_data %>%
  mutate(INSURC22 = case_when(
    INSURC22 %in% c("<65 Any Private", "<65 Public Only") ~ "<65 Any Private/Public",
    INSURC22 %in% c("65+ Medicare and Private", "65+ Medicare and Other Public Only") ~ "65+ Medicare and Any Private/Public",
     INSURC22 %in% c("65+ No Medicare and Any Public/Private") ~ "65+ No Medicare and Any Private/Public",
    TRUE ~ as.character(INSURC22)  # Keep other categories unchanged
  )) %>%
  mutate(INSURC22 = factor(INSURC22))

#Reoder levels
clean_data <- clean_data %>%
  mutate(INSURC22 = factor(INSURC22, levels = c(
    "<65 Uninsured", "<65 Any Private/Public", "65+ Uninsured", "65+ No Medicare and Any Private/Public", "65+ Medicare Only","65+ Medicare and Any Private/Public"
  )))
```

## Logistic Regression Results
### <65 Population
```{r}
#| label: log-reg-under65
#| include: false 

# Separate into 65 under
data_under65 <- clean_data %>% filter(AGE_GROUP != "65+")  # OR AGE_GROUP == "<65"

data_under65_filtered <- data_under65 %>%
  filter(INSURC22 %in% c("<65 Uninsured", "<65 Any Private/Public"))

# Ensure INSURC22 is a factor with "<65 Uninsured" as the reference
data_under65_filtered$INSURC22 <- factor(data_under65_filtered$INSURC22, 
                                         levels = c("<65 Uninsured", "<65 Any Private/Public"))

# Run logistic regression using the filtered dataset
logit_model_under65 <- glm(exposure_psych_drugs ~ INSURC22, data = data_under65_filtered, family = binomial)
```

```{r}
#| label: format-log-reg-under65
#| include: false 
# Extract and format results for APA style
logit_results_under65 <- broom::tidy(logit_model_under65, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    "Insurance Level" = term,
    "OR" = estimate,
    "95% CI (Lower)" = conf.low,
    "95% CI (Upper)" = conf.high,
    "p-value" = p.value
  ) %>%
  mutate(
    OR = sprintf("%.2f", OR),  
    `95% CI (Lower)` = sprintf("%.3f", `95% CI (Lower)`),  
    `95% CI (Upper)` = sprintf("%.3f", `95% CI (Upper)`),  
    `p-value` = case_when(
      `p-value` < 0.001 ~ "< .001",
      TRUE ~ sprintf("%.3f", `p-value`))
  )
```

```{r}
#| label: include-ref-group
#| include: false 
# Convert reference group values to character to match the formatted data and remove intercept
logit_results_under65 <- logit_results_under65 %>%
  filter(!`Insurance Level` %in% c("(Intercept)", "<65 Uninsured")) %>%  # Remove duplicates
  bind_rows(tibble(
    "Insurance Level" = "<65 Uninsured",
    OR = "1.00",
    `95% CI (Lower)` = "—",
    `95% CI (Upper)` = "—",
    `p-value` = "—"
  )) %>%
  arrange(factor(`Insurance Level`, levels = c("<65 Uninsured", "<65 Any Private/Public")))

# Remove variable prefix
logit_results_under65 <- logit_results_under65 %>%
  mutate(`Insurance Level` = gsub("INSURC22", "", `Insurance Level`))  # Remove "INSURC22"
```

```{r}
#| label: tbl-log-reg-under65
#| echo: false 
#| tbl-cap: "Logistic Regression between Insurance Level and Exposure to Psych Drugs with under 65"
#| apa-note: "Reference group is <65 Uninsured (OR = 1.00)"

# Fix APA-style logistic regression table
logit_table_under65 <- flextable(logit_results_under65) %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(j = 2:5, align = "center", part = "body") %>%  # Center-align OR, CI, p-values
  width(j = 2:5, width = 2.5) %>%
  autofit() %>%
  add_footer_lines("Note. Reference group is <65 Uninsured (OR = 1.00).") %>%
  italic(j = 5, part = "body") %>%  # Italicize p-values (APA style)
  align(align = "left", part = "footer")  

# Print table
logit_table_under65
```

@tbl-log-reg-under65 presents logistic regression results for the **under 65 population**. 
`r format_logit_results(logit_model_under65, "INSURC22<65 Any Private/Public", "<65 Any Private/Public")` 

### 65+ Population
```{r}
#| label: regroup-insurance-level
#| include: false 
# Separate into 65+ data
data_65plus <- clean_data %>% filter(str_starts(INSURC22, "65+"))

table(clean_data$INSURC22)
table(data_65plus$INSURC22)
#Divide insurance level to no medicare and medicare 
data_65plus2 <- data_65plus %>%
  mutate(INSURC22 = case_when(INSURC22 %in% c("65+ Uninsured", "65+ No Medicare and Any Private/Public") ~ "65+ without Medicare",
    INSURC22 %in% c("65+ Medicare and Any Private/Public", "65+ Medicare Only") ~ "65+ with Medicare",
    TRUE ~ as.character(INSURC22)  # Keep other categories unchanged
  )) %>%
  mutate(INSURC22 = factor(INSURC22))

table(clean_data$AGE_GROUP)
table(data_65plus2$INSURC22)

#Reoder levels
data_65plus2 <- data_65plus2 %>%
  mutate(INSURC22 = factor(INSURC22, levels = c(
  "65+ without Medicare", "65+ with Medicare"
  )))

table(data_65plus2$INSURC22, data_65plus2$exposure_psych_drugs)

```

```{r}
#| label: data-prep-log-reg
#| include: false 
# Run logistic regression
logit_model3 <- glm(exposure_psych_drugs ~ INSURC22, data = data_65plus2, family = binomial)

data_65plus2

# Extract regression results as a tidy dataframe
logit_results_65plus2 <- tidy(logit_model3, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    "Insurance Level" = term,
    "OR" = estimate,
    "95% CI (Lower)" = conf.low,
    "95% CI (Upper)" = conf.high,
    "p-value" = p.value
  )
```

```{r}
#| label: format-log-reg-65plus
#| include: false 
# Ensure OR, CI, and p-value are characters for consistency
logit_results_65plus2 <- logit_results_65plus2 %>%
  mutate(
    OR = sprintf("%.3f", OR),  # Round OR to 3 decimal places
    `95% CI (Lower)` = sprintf("%.3f", `95% CI (Lower)`),  # Round CI lower bound
    `95% CI (Upper)` = sprintf("%.3f", `95% CI (Upper)`),  # Round CI upper bound
    `p-value` = case_when(  # Format p-values properly
      `p-value` < 0.001 ~ "< .001",
      TRUE ~ sprintf("%.3f", `p-value`))) %>%
  mutate(
    `Insurance Level` = gsub("INSURC22", "", `Insurance Level`)  # Remove variable prefix
  ) %>%
   mutate(
    OR = as.character(OR),
    `95% CI (Lower)` = as.character(`95% CI (Lower)`),
    `95% CI (Upper)` = as.character(`95% CI (Upper)`),
    `p-value` = as.character(`p-value`)
  ) %>%
  bind_rows(tibble(
    "Insurance Level" = "65+ without Medicare",
    OR = "1.00",  # Match format
    `95% CI (Lower)` = "—",  # Use placeholder
    `95% CI (Upper)` = "—",
    `p-value` = "—"
  )) %>%
  arrange(factor(`Insurance Level`, levels = c("65+ without Medicare", "65+ with Medicare")))

# Filter out the (Intercept) row before creating the table
logit_results_65plus2 <- logit_results_65plus2 %>%
  filter(`Insurance Level` != "(Intercept)")
```

```{r}
#| label: tbl-log-reg-65over
#| echo: false 
#| tbl-cap: "Logistic Regression Results for Medicare Presence on Psychiatric Drug Exposure for 65+"
#| apa-note: "Reference group is 65+ Uninsured (OR = 1.00)"

# Fix APA-style logistic regression table
logit_table_65plus2 <- flextable(logit_results_65plus2) %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(j = 2:5, align = "center", part = "body") %>%  # Center-align OR, CI, p-values
  width(j = 2:5, width = 2.5) %>%
  autofit() %>%
  add_footer_lines("Note. Reference group is 65+ Uninsured (OR = 1.00).") %>%
  italic(j = 5, part = "body") %>%  # Italicize p-values
  align(align = "left", part = "footer")

# Print table
logit_table_65plus2
```
@tbl-log-reg-65over presents logistic regression results for the **65 over population**. 
`r format_logit_results(logit_model3, "INSURC2265+ with Medicare", "65+ with Medicare")`

## Psychiatric Drug Exposure by Insurance Level
```{r}
#| label: data-facet-prep
#| include: false

custom_labels <- c(
  "Poor/Negative" = "Poor/Negative",
  "Near Poor" = "Near Poor",
  "Low Income" = "Low Income",
  "Middle Income" = "Middle Income",
  "High Income" = "High Income")

fig_facet <- clean_data %>%
  group_by(POVCAT22, INSURC22, exposure_psych_drugs) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(POVCAT22, INSURC22) %>%
  mutate(prop = count / sum(count))
```

::: {.figure}
```{r}
#| label: fig-facet
#| fig-cap: "Exposure to psychiatric drugs by income level and insurance level"
#| apa-note: "Income levels were cut off by MEPS guidelines." 
#| fig.width: 12
#| fig.height: 8
#| echo: false

fig_facet %>%
ggplot (aes(x = INSURC22, y = prop, fill = exposure_psych_drugs)) +
geom_bar(stat = "identity", position = "fill") +
facet_wrap(~ POVCAT22, 
           nrow = 2, # Set 2 rows for better spacing
            scales = "free_y", # Allow y-axis to be different for each facet
           labeller = labeller(POVCAT22 = custom_labels)) + # Custom facet labels
scale_fill_manual(name = "Exposure to Psychiatric Drugs", values = c("#FFDC5F","#102A71")) + # Renaming legend
labs(x = "Insurance Level", y = "Proportion Exposed") +
theme_minimal() +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),
      axis.text.y = element_text(size=10, margin = margin(r = 10)), 
      strip.text = element_text(size = 10, face = "bold"), 
      axis.title.y = element_text(margin = margin(r = 15), size = 12),  # Add space between y-axis label and plot
      axis.title.x = element_text(margin = margin(t = 10), size = 12))  # Add space between x-axis label and plot 

```
@fig-facet visualizes psychiatric drug exposure proportions by **insurance level** and **income group**. 

Individuals with private/public insurance consistently had higher psychiatric drug exposure across all income levels except for the Near Poor group. Notably, psychiatric drug exposure was lowest among the uninsured, supporting the hypothesis that lack of insurance acts as a barrier to accessing psychiatric treatment.

## Psychiatric Drug Exposure by Age Group
```{r}
#| label: plot2-dataclean
#| include: false 

# Create a new variable for Age Group
clean_data2 <- clean_data %>%
  mutate(Age_Group = ifelse(grepl("^65\\+", INSURC22), "65+", "<65")) %>%
  mutate(INSURC22 = case_when(INSURC22 %in% c("65+ Uninsured", "65+ No Medicare and Any Private/Public") ~ "65+ without Medicare",
    INSURC22 %in% c("65+ Medicare and Any Private/Public", "65+ Medicare Only") ~ "65+ with Medicare",
    TRUE ~ as.character(INSURC22)  # Keep other categories unchanged
  )) %>%
  mutate(INSURC22 = factor(INSURC22))

# Reorder insurance categories as specified
clean_data2 <- clean_data2 %>%
  mutate(
    INSURC22 = factor(INSURC22, levels = c(
      # <65 Group (Left Panel)
      "<65 Uninsured",
      "<65 Any Private/Public",
      # 65+ Group (Right Panel)
      "65+ without Medicare",
      "65+ with Medicare")))
```

```{r}
#| label: fig-psych-insurc
#| fig-cap: "Psychiatric Drug Exposure by Insurance Coverage "
#| apa-note: "Insurance levels were grouped differently due to small sample size and lack of details regarding what type of private or public insurance individual has." 
#| fig.width: 12
#| fig.height: 6
#| echo: false 

# Generate the reordered faceted plot
fig_2 <- ggplot(clean_data2, aes(x = INSURC22, fill = as.factor(exposure_psych_drugs))) +
  geom_bar(position = "fill") +  
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Insurance Level", y = "Percentage",
       fill = "Psychiatric Drug Exposure") +
  scale_fill_manual(name = "Exposure to Psychiatric Drugs", values = c("#c6e6e3", "#cdb4da")) + 
  theme(axis.title.y = element_text(margin = margin(r = 10), size=12), # Add space between y-axis label and plot
    axis.title.x = element_text(margin = margin(r = 15), size=12)) +  # Add space between x-axis label and plot) 
  theme_minimal()
  
print(fig_2)
```
@fig-psych-insurc is a stacked bar plot that visualizes psychiatric drug exposure proportions by **insurance level** using three Insurance Level groups. 

In both age groups **(<65 and 65+)**, the proportion of individuals who have insurance have higher exposure to psychiatric drugs than those who are uninsured. 

## Psychiatric Drug Prescription by Age and Sex

```{r}
#| label: fig-age-sex
#| fig-cap: "Psychiatric Drug Exposure by Age and Sex"
#| fig.width: 14
#| fig.height: 6
#| echo: false 

ggplot(clean_data2, aes(x = Age_Group, y = psych_drug_count, fill = SEX)) +
  geom_violin(alpha = 0.3) +  # Creates density visualization
  scale_y_continuous(breaks = seq(0, max(clean_data2$psych_drug_count, na.rm = TRUE), by = 1)) +  # Set increments to 1
  scale_fill_manual(name = "Sex", values = c("Male" = "#4169E1", "Female" = "#FF9999")) +
  labs(
    x = "Age Group",
    y = "Prescribed Psychiatric Drug Amount"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, vjust = 0.5),  # Adjust x-axis text
    axis.text.y = element_text(size = 10),  # Adjust y-axis text
    axis.title.y = element_text(margin = margin(r = 15), size = 12),  # Add space between y-axis label and plot
    axis.title.x = element_text(margin = margin(r = 15), size = 12)  # Add space between x-axis label and plot
  )
```
@fig-age-sex illustrates the distribution of **psychiatric drug prescriptions** by **age group** and **sex**. 

Consistent with previous findings, **females** appear slightly more likely than males to receive psychiatric medications, although the distribution between age and sex groups seem approximately similar. While most individuals receive **0–2 psychiatric drugs**, a small subset receives a significantly higher number.


## Age and Number of Psychiatric Diagnosis
```{r}
#| label: fig-age-psychcond
#| fig-cap: "Number of Psychiatric Conditions by Age"
#| fig.width: 14
#| fig.height: 6
#| echo: false 

ggplot(clean_data2, aes(x = AGE31X, y = psych_conditions_count, color = SEX)) +  
  geom_point(alpha = 0.8) +  # Scatter plot points
  geom_jitter(width = 0.3, height = 0.2, alpha = 0.6) +  # Avoid point overlap
  geom_smooth(method = "lm", se = TRUE, linetype = "solid", size = 1) +  # Trendline
  scale_color_manual(values = c("Male" = "#8da5ed", "Female" = "#fda172")) +  # Custom colors for sex
  labs(
    x = "Age",
    y = "Number of Diagnosed Psychiatric Conditions",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, vjust = 0.5),  # Adjust x-axis text
    axis.text.y = element_text(size = 10),  # Adjust y-axis text
    axis.title.y = element_text(margin = margin(r = 15), size = 12),  # Add space between y-axis label and plot
    axis.title.x = element_text(margin = margin(r = 15), size = 12)  # Add space between x-axis label and plot
  )
```
@fig-age-psychcond displays the relationship between both **age** and **sex** and the **number of psychiatric diagnoses**. 

Across all ages, psychiatric diagnoses were most frequently **0 or 1**, suggesting that multiple diagnoses were uncommon in this population. The nearly flat regression line suggests that neither sex or age strongly predicts the number of psychiatric conditions.

# Discussion

The findings of this study highlight a strong association between insurance coverage and psychiatric drug exposure, reinforcing the idea that insurance plays a crucial role in mental health treatment access. @tbl-chi-square confirm that all covariates—age, sex, race/ethnicity, poverty level, and insurance status—are significantly related to psychiatric drug exposure, while logistic regression results in @tbl-log-reg-under65 and @tbl-log-reg-65over show that insured individuals, regardless of age or Medicare eligibility, are nearly twice as likely to receive psychiatric prescriptions as their uninsured counterparts. Notably, the "Near Poor" subgroup in @tbl-subgroup-analysis did not reach statistical significance, likely due to small sample size limitations rather than a true absence of association, a limitation that is also evident in @fig-facet, where data sparsity affects certain groups. @fig-psych-insurc visually confirms the logistic regression findings, showing that insured individuals consistently have higher psychiatric drug exposure. Further, @fig-age-sex suggests that while age and sex predict psychiatric drug exposure, they do not strongly influence the number of psychiatric prescriptions a person receives, while @fig-age-psychcond indicates that age and sex do not predict the number of psychiatric conditions a person has. These findings suggest that the association between age and psychiatric drug exposure is likely driven by treatment access rather than a higher prevalence of psychiatric conditions in older individuals. Despite these insights, this study is limited by sample size constraints and the cross-sectional nature of MEPS, which prevents causal inference. Future research should examine longitudinal trends in psychiatric drug access and explore how policy interventions, such as Medicaid expansion and cost-sharing reductions, can further mitigate disparities in psychiatric medication use.
