---
title: "Insurance Coverage and Psychiatric Drug Exposure"
author:
- name: Sunny Lee 
  corresponding: true
  email: snylee@uchicago.edu
  affiliations:
- name: University of Chicago
department: Department of Comparative Human Development
address: 1101 E 58th St
city: Chicago
region: IL
postal-code: 60637
abstract: "This is my abstract."
keywords: [MEPS, Medical Expenditure Panel Survey, Mental Health, Poverty, Prescribed Medicine, Psychiatric Drug, Drug Expenditure]
author-note:
disclosures:
conflict of interest: The author has no conflict of interest to declare.
format:
bibliography: bibliography.bib
apaquarto-html: default
apaquarto-pdf: default
---

# Introduction
Mental health disorders pose a significant global challenge, with more than half of individuals suffering from mental illnesses lacking access to proper care [@pateletal_2010_reducing]. This treatment gap is widespread across nations but disproportionately affects individuals in lower-income brackets. The relationship between poverty, mental health, and social risks—including stress, stigma, and unemployment—has been extensively studied, revealing a cyclical interaction in which mental illness can lead to financial hardship while economic instability exacerbates psychiatric conditions (Knifton & Inglis, 2020). Despite the heightened risk of psychiatric disorders among socioeconomically disadvantaged populations, significant barriers—including financial constraints, lack of insurance, and limited healthcare access—impede treatment utilization (Dong et al., 2021).

One of the primary obstacles to mental healthcare access is the cost of psychiatric medications, which constitute a substantial portion of healthcare expenditures (Dong et al., 2021). The rising cost of psychotropic drugs has placed an increasing financial burden on low-income families, leading to a higher prevalence of unmet healthcare needs (Knifton & Inglis, 2020). Insurance coverage plays a crucial role in mitigating these disparities, as research has shown that expanding coverage improves access to prescription medications. For example, a 2018 study in Oregon found that Medicaid coverage reduced the prevalence of untreated depression by over 60%, while increasing the use of antidepressants and other psychiatric medications (McConnell et al., 2018). Similarly, the Medicare Improvement for Patients and Providers Act of 2008 facilitated greater psychotropic medication use by reducing the cost burden of prescriptions (Polinski et al., 2010). These findings underscore the importance of financial barriers in psychiatric medication adherence, highlighting the role of insurance policies in reducing disparities in mental healthcare access.

Given the critical role of insurance coverage in determining psychiatric drug access, this study examines the following research question: What is the relationship between insurance coverage and exposure to psychiatric drugs? Based on prior research, the study hypothesizes that individuals with comprehensive insurance coverage will have greater exposure to psychiatric medications compared to those with limited or no insurance.

This study utilizes data from the Medical Expenditure Panel Survey (MEPS) to investigate the association between insurance coverage and psychiatric drug exposure, obtained from the Prescribed Medicine File. The primary outcome variable, psychiatric drug exposure, is defined as a binary indicator of whether an individual was prescribed psychiatric medication, as reported by the household and confirmed by a pharmacy with individual consent. The key predictor variable, insurance coverage, is categorized into private insurance, public insurance (Medicaid, Medicare), and uninsured status. Several covariates that may influence psychiatric drug exposure are also considered, including:

Number of psychiatric diagnoses
Demographic characteristics: Age, race/ethnicity, and sex
Socioeconomic status: Poverty/income level
Additionally, to further explore the role of insurance coverage in psychiatric medication access, the study examines three distinct subgroups:

Individuals diagnosed with a psychiatric condition but not prescribed medication
Individuals diagnosed with a psychiatric condition and prescribed medication
Individuals without a psychiatric diagnosis who were prescribed psychiatric medication
By analyzing these subgroups, this study seeks to assess whether insurance status not only influences psychiatric medication access but also diagnostic and treatment patterns. The findings may contribute to healthcare policy discussions, particularly in advocating for expanded insurance coverage and reduced cost-related barriers to psychiatric treatment.

# Methods

## Data Source
This study utilizes data from the **Medical Expenditure Panel Survey (MEPS)**, a nationally representative survey that collects information on healthcare utilization, expenditures, and insurance coverage in the United States. Specifically, the **2022 MEPS Full-Year Consolidated File (hc243), Medical Conditions File (h241), and Prescribed Medicine File (h239a)** were used. The **household component** of MEPS provides person-level data, while the **medical conditions and prescribed medicine files** include condition- and prescription-level data, respectively. The **2022 dataset** included **21,747 individuals**. Participants included in this study were individuals aged **18 years and older**.

## Data Preparation
### Variable Construction
- **Psychiatric Drug Exposure:** A binary outcome variable indicating whether an individual was prescribed and purchased a psychiatric drug. This was derived from the **h239a prescribed medicine file**, where only medications classified under **psychiatric therapeutic classes** were retained.
- **Psychiatric Diagnoses:** Conditions were filtered using **ICD codes** from the **h241 medical conditions file**, and only psychiatric conditions were retained.
- **Insurance Coverage:** The original MEPS insurance categories were **recoded** as follows:
  - **<65 Any Private and <65 Public Only → <65 Any Private/Public**
    - **65+ Medicare and Private/Public, 65+ Medicare and Other Public Only → 65+ Medicare and Any Private/Public**
      - **65+ No Medicare and Any Public/Private → 65+ No Medicare and Any Private/Public**
Age groups were categorized as **<65 years** and **65+ years** to align with Medicare eligibility thresholds.

## Key Variables
- **Outcome Variable:** Psychiatric drug exposure (Yes/No)
- **Independent Variable:** Insurance coverage status, categorized as:
  - **<65 Uninsured**
  - **<65 Any Private/Public**
  - **65+ without Medicare**
  - **65+ with Medicare**
- **Covariates:**
  - Age
  - Sex
  - Race/Ethnicity
  - Income Level

## Statistical Analysis

### Descriptive Analysis
Descriptive statistics were computed for demographic and socioeconomic characteristics, stratified by psychiatric drug exposure status. **Chi-square tests** examined bivariate relationships between psychiatric drug exposure and key variables.

### Logistic Regression Analysis
Separate **logistic regression models** were performed for each age group to assess the association between **insurance coverage and psychiatric drug exposure**. Odds ratios (OR) with **95% confidence intervals (CIs)** were estimated. The reference groups were:
- **<65 Uninsured** for the <65 model
- **65+ without Medicare** for the 65+ model

All analyses were conducted using **R version 2024.12.0+467** with the `tidyverse`, `broom`, and `flextable` packages.

# Results

```{r}
#| label: Setting up the quarto for data cleaning
#| echo: false

library(readr) #Run for every new R session

#Source an R script that has imported ASCII file and data-cleaned, extracting only relevant variables
source ("data-wrangling/Loading and cleaning h243.R")

#Source an R script that merges full year consolidated and medical conditions file 
source ("data-wrangling/Merged h243 and h241.R")

#Source all 
source ("data-wrangling/Added h239a to data.R")

```

```{r}
#| label: Load packages for descriptive statistics and relevant R Script 
#| echo: false 

#install.packages("gtsummary") #commented out the install function so it doesn't install everytime
#install.packages("flextable")
library(gtsummary)
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(scales) # squishing scales (and other outlier handling)
library(psych) # common analyses and visualizations in psychology
library(stats) # basic statistics
library(lme4) # mixed models
library(patchwork) # "stitch" together multiple plots
library(rempsyc) # imperfect APA tables
library(broom) # tidy model output
library(papaja) # general APA friendly stuff, but old and not always quarto-friendly
# load flextable after papaja because they both have a theme_apa() function
# and papaja's one (for plots) isn't very reliable; we want flextable's (for tables) to override ("mask") it
library(flextable) # yet another APA table option


source("functions.R")
```

```{r}
#| label: analysis-preferences
#| echo: false

# Seed for random number generation
set.seed(24)

# Set global chunk options
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed,
  fig.height = 6,
	fig.width = 7,
	message = FALSE,
	warning = FALSE,
	comment = ""
  )
```

## Descriptive Statistics

```{r}
#| label: Descriptive Statistics by exposure to psychiatric drugs 
#| echo: false 

# Generate descriptive statistics by psychiatric drug exposure
table1 <- clean_data %>%
  group_by(exposure_psych_drugs) %>%
  summarise(
    `Age (Mean ± SD)` = paste0(round(mean(AGE31X, na.rm = TRUE), 2), 
                                " (±", round(sd(AGE31X, na.rm = TRUE), 2), ")"),
    `Male (%)` = paste0(round(mean(SEX == "Male", na.rm = TRUE) * 100, 2), "%"),
    `Female (%)` = paste0(round(mean(SEX == "Female", na.rm = TRUE) * 100, 2), "%"),
    `Hispanic (%)` = paste0(round(mean(RACETHX == "Hispanic", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic White (%)` = paste0(round(mean(RACETHX == "Non-Hispanic White", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic Black (%)` = paste0(round(mean(RACETHX == "Non-Hispanic Black", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic Asian (%)` = paste0(round(mean(RACETHX == "Non-Hispanic Asian", na.rm = TRUE) * 100, 2), "%"),
    `Non-Hispanic Other/Multiple (%)` = paste0(round(mean(RACETHX == "Non-Hispanic Other/Multiple", na.rm = TRUE) * 100, 2), "%"),
    `Poor/Negative (%)` = paste0(round(mean(POVCAT22 == "Poor/Negative", na.rm = TRUE) * 100, 2), "%"),
    `Near Poor (%)` = paste0(round(mean(POVCAT22 == "Near Poor", na.rm = TRUE) * 100, 2), "%"),
    `Low Income (%)` = paste0(round(mean(POVCAT22 == "Low Income", na.rm = TRUE) * 100, 2), "%"),
    `Middle Income (%)` = paste0(round(mean(POVCAT22 == "Middle Income", na.rm = TRUE) * 100, 2), "%"),
    `High Income (%)` = paste0(round(mean(POVCAT22 == "High Income", na.rm = TRUE) * 100, 2), "%")
  ) %>%
  pivot_longer(cols = -exposure_psych_drugs, names_to = "Variable", values_to = "Value") %>%
  pivot_wider(names_from = exposure_psych_drugs, values_from = Value)

# Rename columns for better readability
colnames(table1) <- c("Variable", "No Exposure (%)", "Exposed (%)")
```

```{r}
#| label: Descriptive Statistics Table
#| fig-cap: "Descriptive statistics by exposure to psychiatric drugs"
#| echo: true

# Format the table with APA-style formatting
table1 %>%
  kable("html", 
        caption = "Table 1: Descriptive Statistics by Psychiatric Drug Exposure") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  pack_rows("Age", 1, 1) %>%
  pack_rows("Sex", 2, 3) %>%
  pack_rows("Race/Ethnicity", 4, 8) %>%
  pack_rows("Income Level", 9, 13) %>%
  add_indent(c(2:3, 4:8, 9:13))  # Indent subcategories

table1
```
Table 1 presents **descriptive statistics** for the study population by psychiatric drug exposure status. Individuals exposed to psychiatric drugs were more likely to be older, female, and from **higher-income groups**.

```{r}
#|label: Age Categorize
#|echo: false

#Make Age a Categorical Variable 
clean_data <- clean_data %>%
  mutate(AGE_GROUP = as.factor(
      case_when(
        AGE31X >= 18 & AGE31X <= 25 ~ "18-25",
        AGE31X >= 26 & AGE31X <= 45 ~ "26-45",
        AGE31X >= 46 & AGE31X <= 64 ~ "46-64",
        AGE31X >= 65 ~ "65+")))
```

## Chi-Square Test Results
```{r}
#| label: Chi-Square Test Table Preparation
#| echo: false

# Run Chi-Square tests
sex_chi <- chisq.test(table(clean_data$SEX, clean_data$exposure_psych_drugs))
race_chi <- chisq.test(table(clean_data$RACETHX, clean_data$exposure_psych_drugs))
age_chi <- chisq.test(table(clean_data$AGE_GROUP, clean_data$exposure_psych_drugs))  # Use AGE_GROUP instead of Age
pov_chi <- chisq.test(table(clean_data$POVCAT22, clean_data$exposure_psych_drugs))

# Create a data frame with results
chi_square_bygroup <- data.frame(
  Variable = c("Sex", "Race/Ethnicity", "Age Group", "Poverty Level"),
  `X²` = round(c(sex_chi$statistic, race_chi$statistic, age_chi$statistic, pov_chi$statistic), 3),  # Round X² values to 3 decimal places
  df = c(sex_chi$parameter, race_chi$parameter, age_chi$parameter, pov_chi$parameter),
  p_value = c(sex_chi$p.value, race_chi$p.value, age_chi$p.value, pov_chi$p.value)
)

# Format p-values for APA style
chi_square_bygroup$p_value <- as.character(ifelse(chi_square_bygroup$p_value < .001, "< .001", round(chi_square_bygroup$p_value, 3)))

```

```{r}
#| label: Make Table for Chi-Square Test
#| fig-cap: "Relationship between Insurance Level and Exposure to Psychiatric Drugs based on Covariates"
#| echo: true

# Make APA-style Chi-Square Test Table
chi_square_table <- flextable(chi_square_bygroup) %>%
  set_caption("Table 2\nChi-Square Test Results for Psychiatric Drug Exposure") %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(align = "center", part = "body") %>%
  width(j = 2:4, width = 2) %>%  
  fontsize(size = 10, part = "all") %>%
  autofit() %>%
  italic(j = 4, part = "body")  # Italicize p-values (APA style)

# Print the table
chi_square_table
```
Chi-square tests indicated significant differences in **psychiatric drug exposure** across sex, race/ethnicity, age group, and poverty level (Table 2).


## Subgroup Chi-Square Analysis
```{r}
#| label: Subgroup Analysis and n-value 
#| echo: false 

# Make wide table
summary_table <- clean_data %>%
  group_by(POVCAT22, INSURC22, exposure_psych_drugs) %>%
  summarise(count = n(), .groups = "drop") %>% #Drop groups after summarizing 
  tidyr::pivot_wider(names_from = exposure_psych_drugs, values_from = count, values_fill = 0)

# Chi-square test within each income level and Fischer's exact for smaller values  

subgroup_chisq <- clean_data %>%
  group_by(POVCAT22) %>%
  summarise(
    p_value = tryCatch({
        if (any(table(INSURC22, exposure_psych_drugs) < 5)) {
          fisher.test(table(INSURC22, exposure_psych_drugs), simulate.p.value = TRUE)$p.value}          else {
          chisq.test(table(INSURC22, exposure_psych_drugs))$p.value
        }
      },
      error = function(e) NA ))  # If error occurs, return NA

# Calculate "n" values for individuals with "Yes" in exposure to psych drugs per income level
n_values <- clean_data %>%
  filter(exposure_psych_drugs == "Yes") %>%
  group_by(POVCAT22) %>%
  summarise(n = n(), .groups = "drop")

# Merge the "n" values with chi-square results
subgroup_chisq <- subgroup_chisq %>%
  left_join(n_values, by = "POVCAT22")

# Format p-values for APA style
subgroup_chisq <- subgroup_chisq %>%
  mutate(p_value = ifelse(p_value < 0.001, "< .001", sprintf("%.3f", p_value)))

# Rename columns for APA format
subgroup_chisq <- subgroup_chisq %>%
  rename(
    "Income Level" = POVCAT22,
    "Chi-Square / Fisher's p-value" = p_value,
    "n (Exposed to Psych Drugs)" = n
  )
```

```{r}
#| label: Subgroup Analysis and n-value table
#| fig-cap: "Table 3: Relationship between Insurance Level and Exposure to Psychiatric Drugs based on Income Level"
#| echo: true


# Create APA-style table with "n" values
table3 <- subgroup_chisq %>%
  flextable() %>%
  set_caption("Table 3: Subgroup Chi-Square for Income Level") %>%
  theme_vanilla() %>%  # APA-style theme
  bold(part = "header") %>%  # Make headers bold
  align(align = "center", part = "all") %>%  # Center align all content
  fontsize(size = 10, part = "all") %>%
  width(j = 2, width = 2) %>%  # Adjust X² column width
  width(j = 3, width = 2) %>%  # Adjust df column width
  autofit()

print(table3)
```
Table 3 presents subgroup analyses stratified by **income level**. The association between insurance coverage and psychiatric drug exposure remained significant in **all income groups** except the **Near Poor** category.



```{r}
#| label: Changing names for consistency and regrouping of insurance levels 
#| echo: false 
clean_data <- clean_data %>%
  mutate(INSURC22 = case_when(
    INSURC22 %in% c("<65 Any Private", "<65 Public Only") ~ "<65 Any Private/Public",
    INSURC22 %in% c("65+ Medicare and Private", "65+ Medicare and Other Public Only") ~ "65+ Medicare and Any Private/Public",
     INSURC22 %in% c("65+ No Medicare and Any Public/Private") ~ "65+ No Medicare and Any Private/Public",
    TRUE ~ as.character(INSURC22)  # Keep other categories unchanged
  )) %>%
  mutate(INSURC22 = factor(INSURC22))


#Reoder levels
clean_data <- clean_data %>%
  mutate(INSURC22 = factor(INSURC22, levels = c(
    "<65 Uninsured", "<65 Any Private/Public", "65+ Uninsured", "65+ No Medicare and Any Private/Public", "65+ Medicare Only","65+ Medicare and Any Private/Public"
  )))


```

## Logistic Regression Results
### <65 Population
```{r}
#| label: Logistic Regression between Insurance Level and Exposure to Psych Drugs with <65 Data Preparation
#| echo: false 

# Separate into 65 under
data_under65 <- clean_data %>% filter(AGE_GROUP != "65+")  # OR AGE_GROUP == "<65"

data_under65_filtered <- data_under65 %>%
  filter(INSURC22 %in% c("<65 Uninsured", "<65 Any Private/Public"))

# Ensure INSURC22 is a factor with "<65 Uninsured" as the reference
data_under65_filtered$INSURC22 <- factor(data_under65_filtered$INSURC22, 
                                         levels = c("<65 Uninsured", "<65 Any Private/Public"))

# Run logistic regression using the filtered dataset
logit_model_under65 <- glm(exposure_psych_drugs ~ INSURC22, data = data_under65_filtered, family = binomial)

# Extract and format results for APA style
logit_results_under65 <- broom::tidy(logit_model_under65, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    "Insurance Level" = term,
    "OR" = estimate,
    "95% CI (Lower)" = conf.low,
    "95% CI (Upper)" = conf.high,
    "p-value" = p.value
  ) %>%
  mutate(
    OR = sprintf("%.2f", OR),  
    `95% CI (Lower)` = sprintf("%.3f", `95% CI (Lower)`),  
    `95% CI (Upper)` = sprintf("%.3f", `95% CI (Upper)`),  
    `p-value` = case_when(
      `p-value` < 0.001 ~ "< .001",
      TRUE ~ sprintf("%.3f", `p-value`))
  )

# Convert reference group values to character to match the formatted data and remove intercept
logit_results_under65 <- logit_results_under65 %>%
  filter(!`Insurance Level` %in% c("(Intercept)", "<65 Uninsured")) %>%  # Remove duplicates
  bind_rows(tibble(
    "Insurance Level" = "<65 Uninsured",
    OR = "1.00",
    `95% CI (Lower)` = "—",
    `95% CI (Upper)` = "—",
    `p-value` = "—"
  )) %>%
  arrange(factor(`Insurance Level`, levels = c("<65 Uninsured", "<65 Any Private/Public")))

# Remove variable prefix
logit_results_under65 <- logit_results_under65 %>%
  mutate(`Insurance Level` = gsub("INSURC22", "", `Insurance Level`))  # Remove "INSURC22"

```

```{r}
#| label: Logistic Regression between Insurance Level and Exposure to Psych Drugs with <65
#| echo: true 
#| fig-cap: "Logistic Regression between Insurance Level and Exposure to Psych Drugs with <65

# Fix APA-style logistic regression table
logit_table_under65 <- flextable(logit_results_under65) %>%
  set_caption("Table 4\nLogistic Regression Results for <65 Insurance and Psychiatric Drug Exposure") %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(j = 2:5, align = "center", part = "body") %>%  # Center-align OR, CI, p-values
  width(j = 2:5, width = 2.5) %>%
  autofit() %>%
  italic(j = 5, part = "body") %>%  # Italicize p-values (APA style)
  add_footer_lines("Note: Reference group is <65 Uninsured (OR = 1.00).") %>%
  align(align = "left", part = "footer")  

# Print table
logit_table_under65
```
Table 4 presents logistic regression results for the **<65 population**. Individuals with **any private/public insurance** had **2.31 times higher odds** of psychiatric drug exposure compared to **uninsured individuals (p < .001).**

### 65+ Population
```{r}
#| label: Regrouping for Logistic Regression Results for Insurance Level (Two groups) and Psychiatric Drug Exposure for 65+ Data Preparation
#| echo: false 

# Separate into 65+ data
data_65plus <- clean_data %>% filter(AGE_GROUP == "65+")

#Divide insurance level to no medicare and medicare 
data_65plus2 <- data_65plus %>%
  mutate(INSURC22 = case_when(INSURC22 %in% c("65+ Uninsured", "65+ No Medicare and Any Private/Public") ~ "65+ without Medicare",
    INSURC22 %in% c("65+ Medicare and Private/Public", "65+ Medicare Only") ~ "65+ with Medicare",
    TRUE ~ as.character(INSURC22)  # Keep other categories unchanged
  )) %>%
  mutate(INSURC22 = factor(INSURC22))

#Reoder levels
data_65plus2 <- data_65plus2 %>%
  mutate(INSURC22 = factor(INSURC22, levels = c(
  "65+ without Medicare", "65+ with Medicare"
  )))
```

```{r}
#| label: Logistic Regression Results for Insurance Level (Two groups) and Psychiatric Drug Exposure for 65+ Data Preparation
#| echo: false 
# Run logistic regression
logit_model3 <- glm(exposure_psych_drugs ~ INSURC22, data = data_65plus2, family = binomial)

# Extract regression results as a tidy dataframe
logit_results_65plus2 <- tidy(logit_model3, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    "Insurance Level" = term,
    "OR" = estimate,
    "95% CI (Lower)" = conf.low,
    "95% CI (Upper)" = conf.high,
    "p-value" = p.value
  )

# Remove intercept row and clean insurance level names
logit_results_65plus2 <- logit_results_65plus2 %>%
  filter(`Insurance Level` != "(Intercept)") %>%
  mutate(
    `Insurance Level` = gsub("INSURC22", "", `Insurance Level`)  # Remove variable prefix
  )



# Ensure OR, CI, and p-value are characters for consistency
logit_results_65plus2 <- logit_results_65plus2 %>%
  mutate(
    OR = sprintf("%.3f", OR),  # Round OR to 3 decimal places
    `95% CI (Lower)` = sprintf("%.3f", `95% CI (Lower)`),  # Round CI lower bound
    `95% CI (Upper)` = sprintf("%.3f", `95% CI (Upper)`),  # Round CI upper bound
    `p-value` = case_when(  # Format p-values properly
      `p-value` < 0.001 ~ "< .001",
      TRUE ~ sprintf("%.3f", `p-value`)
    )
  ) %>%
   mutate(
    OR = as.character(OR),
    `95% CI (Lower)` = as.character(`95% CI (Lower)`),
    `95% CI (Upper)` = as.character(`95% CI (Upper)`),
    `p-value` = as.character(`p-value`)
  ) %>%
  bind_rows(tibble(
    "Insurance Level" = "65+ without Medicare",
    OR = "1.00",  # Match format
    `95% CI (Lower)` = "—",  # Use placeholder
    `95% CI (Upper)` = "—",
    `p-value` = "—"
  )) %>%
  arrange(factor(`Insurance Level`, levels = c("65+ without Medicare", "65+ with Medicare")))

```

```{r}
#| label: Logistic Regression Results for Insurance Level (Two groups) and Psychiatric Drug Exposure for 65+
#| echo: true 
#| fig-cap: "Logistic Regression Results for Medicare Presence on Psychiatric Drug Exposure for 65+"

# Fix APA-style logistic regression table
logit_table_65plus2 <- flextable(logit_results_65plus2) %>%
  set_caption("Table 5\nLogistic Regression Results for 65+ Insurance and Psychiatric Drug Exposure") %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(j = 2:5, align = "center", part = "body") %>%  # Center-align OR, CI, p-values
  width(j = 2:5, width = 2.5) %>%
  autofit() %>%
  italic(j = 5, part = "body") %>%  # Italicize p-values
  add_footer_lines("Note: Reference group is 65+ without Medicare (OR = 1.00).") %>%
  align(align = "left", part = "footer")  

# Print table
logit_table_65plus2
```
Table 5 shows that **65+ individuals with Medicare** had **3.89 times higher odds** of psychiatric drug exposure compared to those **without Medicare (p = .004).**

## Psychiatric Drug Exposure by Insurance Level
```{r}
#| label: Exposure to psychiatric drugs by income level and insurance level 
#| fig-cap: "Exposure to psychiatric drugs by income level and insurance level"
#| echo: true

custom_labels <- c(
  "Poor/Negative" = "Poor/Negative",
  "Near Poor" = "Near Poor",
  "Low Income" = "Low Income",
  "Middle Income" = "Middle Income",
  "High Income" = "High Income")

clean_data %>%
  group_by(POVCAT22, INSURC22, exposure_psych_drugs) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(POVCAT22, INSURC22) %>%
  mutate(prop = count / sum(count)) %>%
  ggplot(aes(x = INSURC22, y = prop, fill = exposure_psych_drugs)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ POVCAT22, 
             nrow = 2, # Set 2 rows for better spacing
             scales = "free_y", # Allow y-axis to be different for each facet
             labeller = labeller(POVCAT22 = custom_labels)) + # Custom facet labels
  scale_fill_manual(name = "Exposure to Psychiatric Drugs", values = c("red", "blue")) + # Renaming legend
   labs(title = "Figure 1: Psychiatric Drug Exposure by Insurance Level Across Income Groups",
       x = "Insurance Level", y = "Proportion Exposed", caption = "Note: Proportions are calculated within each income level.") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        strip.text = element_text(size = 10, face = "bold"), 
        plot.caption = element_text(size = 9, hjust = 0)
        )  # Make facet labels bold
```

## Psychiatric Drug Exposure by Age Group
```{r}
#| label: Plot of Psychiatric Drug Exposure by Insurance Coverage Data Cleaning 
#| echo: false 

# Create a new variable for Age Group
clean_data2 <- clean_data %>%
  mutate(Age_Group = ifelse(grepl("^65\\+", INSURC22), "65+", "<65")) %>%
  mutate(INSURC22 = case_when(INSURC22 %in% c("65+ Uninsured", "65+ No Medicare and Any Private/Public") ~ "65+ without Medicare",
    INSURC22 %in% c("65+ Medicare and Any Private/Public", "65+ Medicare Only") ~ "65+ with Medicare",
    TRUE ~ as.character(INSURC22)  # Keep other categories unchanged
  )) %>%
  mutate(INSURC22 = factor(INSURC22))

# Reorder insurance categories as specified
clean_data2 <- clean_data2 %>%
  mutate(
    INSURC22 = factor(INSURC22, levels = c(
      # <65 Group (Left Panel)
      "<65 Uninsured",
      "<65 Any Private/Public",
      
      # 65+ Group (Right Panel)
      "65+ without Medicare",
      "65+ with Medicare"))
  )

```

```{r}
#| label: Psychiatric Drug Exposure by Insurance Coverage 
#| fig-cap: "Psychiatric Drug Exposure by Insurance Coverage "
#| echo: true 

# Generate the reordered faceted plot
ggplot(clean_data2, aes(x = INSURC22, fill = as.factor(exposure_psych_drugs))) +
  geom_bar(position = "fill") +  # Normalized proportions
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Figure 2: Psychiatric Drug Exposure by Insurance Coverage and Age Group",
       x = "Insurance Coverage",
       y = "Percentage",
       fill = "Psychiatric Drug Exposure") +
  scale_fill_manual(name = "Exposure to Psychiatric Drugs", values = c("lightpink", "lightblue")) + # Renaming legend
  theme_minimal() +
  coord_flip() +
  facet_wrap(~ Age_Group, scales = "free_y")  # Facet by Age Group
```

Needed to group differently due to small sample size and limitation of lack of details regarding what private/public insurance individual has. 
